Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-06 Documents/Paper/ms.tex
Working on intro

\documentclass[letterpaper, 11pt]{article}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[left=2.5cm,top=2.5cm,right=2.5cm]{geometry}
\usepackage{parskip}
\usepackage{amsmath}

@@

In addition to these well-known challenges,
{\it phase transitions} \citep{skilling} can also cause difficulties for
Bayesian computation. Phase transitions occur when the posterior distribution
is a mixture of a ``slab'' with high volume but low density, and a ``spike''
with low volume but high density. To understand why this causes problems,
imagine sampling the posterior using the Metropolis algorithm. If the sampler
is in the slab, it will only rarely propose a move into the spike, because the
spike's volume is so small. If the sampler is in the spike, a proposal to move
into the slab will have a very low acceptance probability because the density
of the slab is so low relative to that of the spike.

@@

Diffusive Nested Sampling
\citep[DNS][]{dnest} replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained versions of the prior, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of annealed importance sampling
\citep{neal} or parallel tempering \citep{pt},
where the target distribution is modified from the posterior to something
``easier''. However,
the sequence of distributions used in Nested Sampling avoids some of the
problems with annealing-based methods. In particular, there is no need to choose
a temperature schedule (a potentially large set of tuning parameters),
and the method does not fail when phase transitions
are present.
Figure~\ref{fig:challenges} shows a mock-up of a two dimensional posterior
density with three modes, one of which contains a strong dependence, and
another of which contains a phase transition.

>>>>

\documentclass[letterpaper, 11pt]{article}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[left=2.5cm,top=2.5cm,right=2.5cm]{geometry}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage{dsfont}

@@

\subsection{Phase Transitions and Nested Sampling}
In addition to the well-known challenges caused by multiple modes and strong
dependencies,
{\it phase transitions} \citep{skilling} can also cause difficulties for
Bayesian computation. Phase transitions occur when the posterior distribution
is a mixture of a ``slab'' with high volume but low density, and a ``spike''
with low volume but high density. To understand why this causes problems,
imagine sampling the posterior using the Metropolis algorithm. If the sampler
is in the slab, it will only rarely propose a move into the spike, because the
spike's volume is so small. If the sampler is in the spike, a proposal to move
into the slab will have a very low acceptance probability because the density
of the slab is so low relative to that of the spike. If the spike has a volume
$V_{\rm spike}$ and the slab has a volume $V_{\rm slab}$ then the number of
MCMC iterations needed to jump from the spike to the slab, or vice versa, is
of order $V_{\rm slab}/V_{\rm spike}$ which may be very large.

In many problems, the posterior distribution is a mixture of a spike component
and a slab component, however the total posterior probability in the slab
component is negligible. In this situation, posterior sampling is unaffected,
however problems can arise when calculating the marginal likelihood using
an annealing-type approach. For example, consider the thermodynamic integral,
which gives the log of the marginal likelihood:

\begin{eqnarray}
\log(\mathcal{Z}) &=& \int_0^1
\left<\log\left[L(\theta)\right]\right>_\beta \,d\beta
\end{eqnarray}

where $L(\theta)$ is the likelihood function, and the expectation is taken
with respect to the ``power posterior'' distributions which are intermediate
between the prior $\pi(\theta)$ and the posterior:

\begin{eqnarray}
p_\beta(\theta) \propto \pi(\theta)L(\theta)^\beta.
\end{eqnarray}

If the posterior distribution is a mixture of a slab and a spike,
then for some range of inverse temperatures $\beta$ the probability mass in
the slab and the spike will be comparable. At these temperatures the MCMC
method will be unable to mix and will give an incorrect estimate of
$\left<\log\left[L(\theta)\right]\right>_\beta$.

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$.

\subsection{Diffusive Nested Sampling}
Diffusive Nested Sampling
\citep[DNS][]{dnest} replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained versions of the prior, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of annealed importance sampling
\citep{neal} or parallel tempering \citep{pt},
where the target distribution is modified from the posterior to something
``easier''. However,
the sequence of distributions used in Nested Sampling avoids some of the
problems with annealing-based methods. In particular, there is no need to choose
a temperature schedule (a potentially large set of tuning parameters),
and the method does not fail when phase transitions
are present.
Figure~\ref{fig:challenges} shows a mock-up of a two dimensional posterior
density with three modes, one of which contains a strong dependence, and
another of which contains a phase transition.

@@
